SU($n$) spin models near the Heisenberg critical point

### 2020-03-27 collective_sun/sun_perturbation.tex
fixed up some text in the shell model section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Breaking SU($n$) symmetry (perturbatively)}
\label{eq:pert_theory}

@@

The second order effective Hamiltonian $H_{\t{eff}}^{(2)}$ is not as
straightforward to simplify due to the presence of a projector
$\P_\Delta$ onto the manifold $\E_\Delta$ of states with excitation
energy $\Delta$.  For each term $X_{\m w}$ of $\hat\O$, this projector
essentially picks off the part of $X_{\m w}$ that is strictly
off-diagonal with respect to the ground- and excited-state manifolds
$\E_0$ and $\E_\Delta$.  To decompose $X_{\m w}$ into components that
generate states of definite excitation energy when acting on PS states
$\M_0$, we pick an arbitrary state $\ket\psi\in\M_0$ and expand (see
Appendix \ref{sec:eigenstates})
\begin{align}
  H_0 X_{\m w} \ket\psi
  = E_0 X_{\m w} \ket\psi + X_{\hat{\m h}\p{\m w}} \ket\psi,
  \label{eq:diagnosis}
\end{align}
where $\hat{\m h}$ is a linear map on the space of dimension-$M$
tensors, determined entirely by the couplings $h_{pq}$ in $H_0$.  The
image $\hat{\m h}\p{\m w}$ of $\m w$ under $\hat{\m h}$ is then itself
a dimension-$M$ tensor that defines the $M$-body operator
$X_{\hat{\m h}\p{\m w}}$.  To write out $\hat{\m h}\p{\m w}$
explicitly, we first define a matrix
$\m h\equiv\sum_{p,q\in\ZZ_N} h_{pq} \op{p}{q}$ of all couplings in
$H_0$ by enforcing that $h_{pq}=h_{qp}$ and $h_{pp}=0$.  We denote a
contraction of $\m h$ with the $a$-th index of $\m w$ by
$\m h_a\p{\m w}$, such that the $k$-th entry of $\m h_a\p{\m w}$ is
\begin{align}
  \m h_a\p{\m w}_k
  \equiv \sum_{\substack{p\in\ZZ_N\\p\notin k}}
  h_{pk_a} w_{k_1 k_2 \cdots k_{a-1} p k_{a+1} \cdots k_M}
  = \sum_{\substack{p\in\ZZ_N\\p\notin k}} h_{pk_a} w_{k_{a:p}},
\end{align}
where if $k=\p{k_1,k_2,\cdots,k_M}$, then $k_{a:p}$ is the same list
but with the $a$-th entry $k_a$ replaced by $p$.  We similarly define
a tensor $\m h_{\t{P}}\p{\m w}$ with entries
\begin{align}
  \m h_{\t{P}}\p{\m w}_k
  \equiv \sum_{\set{a,b}\in\C_M\p{2}} h_{k_ak_b} w_{k_{a\lra b}},
\end{align}
where $k_{a\lra b}$ is the same as $k$ but with the placement of $k_a$
and $k_b$ swapped.  Finally, we define a tensor $\m h_{\t{D}}\p{\m w}$
with entries
\begin{align}
  \m h_{\t{D}}\p{\m w}_k \equiv h^{\t{D}}_k w_k,
  &&
  h^{\t{D}}_k \equiv \sum_{p\in k} h_p
  - \sum_{\set{a,b}\in\C_M\p{2}} h_{k_ak_b},
  &&
  h_p \equiv \sum_{q\in\ZZ_N} h_{pq}.
\end{align}
These definitions allow us to expand
\begin{align}
  \hat{\m h}\p{\m w}
  \equiv \sum_{a\in\ZZ_M} \m h_a\p{\m w}
  + \m h_{\t{P}}\p{\m w} - \m h_{\t{D}}\p{\m w}.
  \label{eq:multi_body_op}
\end{align}
Viewed an operator on the vector space of dimension-$M$ tensors, the
map $\hat{\m h}$ is real and symmetric under transposition, and is
therefore diagonalizable.  Diagonalizing this operator yields
eigenvalues $\Delta$ and eigenvectors $\m u_\Delta$ for which
$\hat{\m h}\p{\m u_\Delta}=\Delta\m u_\Delta$.  Finding eigenvalues
and eigenvectors of $\hat{\m h}$ nominally requires diagonalizing a
matrix of dimensions $\abs{\D_N\p{M}}\times\abs{\D_N\p{M}}$, but
spatial symmetries of $H_0$ and $X_{\m w}$ such as translational
invariance or isotropy can drastically reduce the complexity of this
eigenvalue problem.  We discuss these reductions in Appendix
\ref{sec:symmetries}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Beyond ground-state perturbation theory}
\label{eq:excited_states}

@@

Denoting all assignments of $N$ (identical) spins to $n$ (distinct)
states by $\A_n\p{N}$, a suitable basis for $\M_0$ is the set of PS
states $\set{\ket{m}:m\in\A_n\p{N}}$, where each
$m=\p{m_1,m_2,\cdots,m_n}$ specifies the occupation number $m_\mu$ of
the single-spin state $\mu$.  For consistency with the expression of
matrix elements in \eqref{eq:matrix_element}, for each
$\ket{m}\in\M_0$ we define $\ket{\1,1,m}\equiv\ket{m}$, with $\1$ the
identity operator, and we define $\1_1\equiv\1$.  We now need to
determine the set of $r$-body operators $\set{O}$ and dimension-$r$
tensors $\set{\m u}$ that define our basis for $\M_r$ with
$r=1,2,\cdots,r_{\t{max}}$.  As an initial guess for $\set{\m u}$, we
can choose the set $\set{\m u_\Delta}_{\m h,r}$ of dimension-$r$
tensors $\m u_\Delta$ that generate eigenstates of $H_0$, i.e.~for
which $\hat{\m h}\p{\m u_\Delta} = \Delta \m u_\Delta$, with
$\hat{\m h}$ defined in \eqref{eq:multi_body_op}.  The set of $r$-body
operators to choose is less obvious, but can be determined by an
exhaustive search over all $r$-body operators relevant to a particular
scenario.  Specifically, the purpose of projecting operators onto
$\col\M_{r_{\t{max}}}$ is to simulate the dynamics induced by some
operator $\hat\O$ of the form in \eqref{eq:pert}, constructed from a
set $\O=\set{\p{X,\m w}}$ containing multi-spin operators $X$.  In
order to find a basis for the {\it relevant} states in $\M_r$, one can
simply consider the set $\set{O}_{\O,r}$ all $r$-spin operators that
can be constructed from products of the multi-spin operators $X$ in
$\O$.

In general, the states $\ket{O,\m u_\Delta,m}$ for all
$\p{O,\m u_\Delta,m}\in\set{O}_{\O,r}\times\set{\m u_\Delta}_{\m h,r}
\times\A_n\p{N}$ will form an overcomplete basis for the relevant
states in $\col\M_r$.  Our last step is therefore to reduce this set
of states to a minimal basis $\B_r$ for $\M_r$.  To this end, we
construct each of $\B_1,\B_2,\cdots,\B_{r_{\t{max}}}$ in sequence,
such that when constructing $\B_r$ we already have a minimal basis
$\col\B_{r-1}\equiv\bigcup_{s<r-1}\B_s$ for $\col\M_{r-1}$.  To
construct $\B_r$, we then consider each
$\p{O,\m u_\Delta,m}\in\set{O}_{\O,r}\times\set{\m u_\Delta}_{\m h,r}
\times\A_n\p{N}$ one at a time.  If $\ket{O,\m u_\Delta,m}$ is
orthogonal to all states in $\col\B_{r-1}$, as well as all other
states we have added to $\B_r$ thus far, then we include it in $\B_r$.
Otherwise, we discard $\ket{O,\m u_\Delta,m}$.  There are two
considerations to keep in mind for this construction: first, one
should always throw out null states $\ket{O,\m u_\Delta,m}=0$, which
can be identified by a vanishing self-overlap,
$\bk{O,\m u_\Delta,m|O,\m u_\Delta,m}=0$.  Second, computing overlaps
to check orthogonality can be computationally intensive.  When
considering a candidate state $\ket{O,\m u_\Delta,m}$ to add to
$\B_r$, one can therefore restrict orthogonality checks to other
states $\ket{Q,\m v_\epsilon,\ell}$ with $\epsilon=\Delta$; if the
excitation energies $\epsilon\ne\Delta$, then these states are
guaranteed to be orthogonal.  Any two states with different occupation
numbers of single-spin states $\mu\in\ZZ_n$ are similarly guaranteed
to be orthogonal.

\vspace{3cm}

\red{[TODO: finish section]}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Breaking SU($n$) symmetry}

@@

The second order effective Hamiltonian $H_{\t{eff}}^{(2)}$ is not as
straightforward to simplify due to the presence of a projector
$\P_\Delta$ onto the manifold $\E_\Delta$ of states with excitation
energy $\Delta$.  For each term $X_{\m w}$ of $\hat\O$, this projector
essentially picks off the part of $X_{\m w}$ that is strictly
off-diagonal with respect to the ground- and excited-state manifolds
$\E_0$ and $\E_\Delta$.  To decompose $X_{\m w}$ into components that
generate states of definite excitation energy when acting on PS states
$\M_0$, we pick an arbitrary state $\ket\psi\in\M_0$ and expand (see
Appendix \ref{sec:eigenstates})
\begin{align}
  H_0 X_{\m w} \ket\psi
  = E_0 X_{\m w} \ket\psi + X_{\hat{\m h}\p{\m w}} \ket\psi,
  \label{eq:diagnosis}
\end{align}
where $\hat{\m h}$ is a linear map on the space of dimension-$M$
tensors, determined entirely by the couplings $h_{pq}$ in $H_0$.  The
image $\hat{\m h}\p{\m w}$ of $\m w$ under $\hat{\m h}$ is then itself
a dimension-$M$ tensor that defines the $M$-body operator
$X_{\hat{\m h}\p{\m w}}$.  To write out $\hat{\m h}\p{\m w}$
explicitly, we first define a matrix
$\m h\equiv\sum_{p,q\in\ZZ_N} h_{pq} \op{p}{q}$ of all couplings in
$H_0$ by enforcing that $h_{pq}=h_{qp}$ and $h_{pp}=0$.  We denote a
contraction of $\m h$ with the $a$-th index of $\m w$ by
$\m h_a\p{\m w}$, such that the $k$-th entry of $\m h_a\p{\m w}$ is
\begin{align}
  \m h_a\p{\m w}_k
  \equiv \sum_{\substack{p\in\ZZ_N\\p\notin k}}
  h_{pk_a} w_{k_1 k_2 \cdots k_{a-1} p k_{a+1} \cdots k_M}
  = \sum_{\substack{p\in\ZZ_N\\p\notin k}} h_{pk_a} w_{k_{a:p}},
\end{align}
where if $k=\p{k_1,k_2,\cdots,k_M}$, then $k_{a:p}$ is the same list
but with the $a$-th entry $k_a$ replaced by $p$.  We similarly define
a tensor $\m h_{\t{P}}\p{\m w}$ with entries
\begin{align}
  \m h_{\t{P}}\p{\m w}_k
  \equiv \sum_{\set{a,b}\in\C_M\p{2}} h_{k_ak_b} w_{k_{a\lra b}},
\end{align}
where $k_{a\lra b}$ is the same as $k$ but with the placement of $k_a$
and $k_b$ swapped.  Finally, we define a tensor $\m h_{\t{D}}\p{\m w}$
with entries
\begin{align}
  \m h_{\t{D}}\p{\m w}_k \equiv h^{\t{D}}_k w_k,
  &&
  h^{\t{D}}_k \equiv \sum_{p\in k} h_p
  - \sum_{\set{a,b}\in\C_M\p{2}} h_{k_ak_b},
  &&
  h_p \equiv \sum_{q\in\ZZ_N} h_{pq}.
\end{align}
These definitions allow us to expand
\begin{align}
  \hat{\m h}\p{\m w}
  \equiv \sum_{a\in\ZZ_M} \m h_a\p{\m w}
  + \m h_{\t{P}}\p{\m w} - \m h_{\t{D}}\p{\m w}.
  \label{eq:multi_body_op}
\end{align}
Viewed an operator on the vector space of dimension-$M$ tensors, the
map $\hat{\m h}$ is real and symmetric under transposition, and is
therefore diagonalizable.  Diagonalizing this operator yields
eigenvalues $\Delta$ and eigenvectors $\m u_\Delta$ for which
$\hat{\m h}\p{\m u_\Delta}=\Delta\m u_\Delta$.  Finding eigenvalues
and eigenvectors of $\hat{\m h}$ nominally requires diagonalizing a
matrix of dimensions $\abs{\D_N\p{M}}\times\abs{\D_N\p{M}}$, but
spatial symmetries of $H_0$ and $X_{\m w}$, such as translational
invariance or isotropy, can drastically reduce the complexity of this
eigenvalue problem.  We discuss these reductions in Appendix
\ref{sec:symmetries}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Beyond ground-state perturbation theory}

@@

Denoting all assignments of $N$ (identical) spins to $n$ (distinct)
states by $\A_n\p{N}$, a suitable basis for $\M_0$ is the set of PS
states $\set{\ket{m}:m\in\A_n\p{N}}$, where $m=\p{m_1,m_2,\cdots,m_n}$
specifies the occupation number $m_\mu$ of each single-spin state
$\mu$.  For consistency with the expression of matrix elements in
\eqref{eq:matrix_element}, for each $\ket{m}\in\M_0$ we define
$\ket{\1,1,m}\equiv\ket{m}$, with $\1$ the identity operator, and we
define $\1_1\equiv\1$.  We now need to determine the set of $r$-body
operators $\set{O}$ and dimension-$r$ tensors $\set{\m u}$ that define
our basis for each $\M_r$.  As an initial guess for $\set{\m u}$, we
can choose the set $\set{\m u_\Delta}_{\m h,r}$ of dimension-$r$
tensors $\m u_\Delta$ that generate eigenstates of $H_0$, i.e.~for
which $\hat{\m h}\p{\m u_\Delta} = \Delta \m u_\Delta$, with
$\hat{\m h}$ defined in \eqref{eq:multi_body_op}.  The set of $r$-body
operators to choose is less obvious, but can be determined by an
exhaustive search over all $r$-body operators relevant to a particular
scenario.  Specifically, the purpose of projecting operators onto
$\col\M_{r_{\t{max}}}$ is to simulate the dynamics induced by some
operator $\hat\O$ of the form in \eqref{eq:pert}, constructed from a
set $\O=\set{\p{X,\m w}}$ containing multi-spin operators $X$.  In
order to find a basis for the {\it relevant} states in $\M_r$, one can
simply consider the set $\set{O}_{\O,r}$ all $r$-spin operators that
can be constructed from products of the multi-spin operators $X$ in
$\O$.

In general, the states $\ket{O,\m u_\Delta,m}$ for all
$\p{O,\m u_\Delta,m}\in\set{O}_{\O,r}\times\set{\m u_\Delta}_{\m h,r}
\times\A_n\p{N}$ will form an overcomplete basis for the relevant
states in $\col\M_r$.  Our last step is therefore to reduce this set
of states to a minimal basis $\B_r$ for $\M_r$.  To this end, we
construct each of $\B_1,\B_2,\cdots,\B_{r_{\t{max}}}$ in sequence,
such that when constructing $\B_r$ we already have a minimal basis
$\col\B_{r-1}\equiv\bigcup_{s<r}\B_s$ for $\col\M_{r-1}$.  To
construct $\B_r$, we then consider each
$\p{O,\m u_\Delta,m}\in\set{O}_{\O,r}\times\set{\m u_\Delta}_{\m h,r}
\times\A_n\p{N}$ one at a time.  If $\ket{O,\m u_\Delta,m}$ is
orthogonal to all states in $\col\B_{r-1}$, as well as all other
states we have added to $\B_r$ thus far, then we include it in $\B_r$.
Otherwise, we discard $\ket{O,\m u_\Delta,m}$.  There are two
considerations to keep in mind for this construction: first, one
should always throw out null states $\ket{O,\m u_\Delta,m}=0$, which
can be identified by a vanishing self-overlap,
$\bk{O,\m u_\Delta,m|O,\m u_\Delta,m}=0$.  Second, computing overlaps
to check for orthogonality can be computationally intensive.  When
considering a candidate state $\ket{O,\m u_\Delta,m}$ to add to
$\B_r$, one can therefore restrict orthogonality checks to other
states $\ket{Q,\m v_\epsilon,\ell}$ with $\epsilon=\Delta$; if the
excitation energies $\epsilon\ne\Delta$, then these states are
guaranteed to be orthogonal.  Any two states with different occupation
numbers of single-spin states $\mu\in\ZZ_n$ are similarly guaranteed
to be orthogonal.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The XXZ model}

\red{[TODO: write section]}
